|

### 2019-08-16 article/article.Rmd
small edits, again

One of the key determinants of efficient public services provision is a robust accountability system [@besley2003incentives; @cameron2004public; @ferejohn1986incumbent; @o1998horizontal]. In its standard definition, accountability is understood as the process of holding authorities responsible for their actions [@finer1941administrative; @mulgan2000accountability; @o1990bureaucratic]. Past studies show that accountability has a positive impact on governance, as it ensures that politicians act on behalf of voters [@freire2010ngp; @moncrieffe1998reconceptualizing], reduces the opportunities for rent-seeking and corruption [@deininger2005does; @wenar2006accountability] and improves the efficiency of the public sector [@adsera2003you; @bjorkman2009power]. Recent research also suggests that accountability leads to higher economic growth because it limits state discretion in the economy and increases long-term investments in human capital [@benhabib2010economic; @suebvises2018social; @ponzetto2018social].

@@

We use the TDP app to conduct two experimental interventions and test its impact on six outcomes related to school completion rates and complaints to public authorities. Overall, our results show that providing information to citizens has very limited impact on policy outcomes. We find that the TDP app increased the likelihood of school completion by 7 percent in the first experiment, but the effect is null in the second intervention. We also find that the app increased the percentage of the executed budget by 157.5 percent in our second experiment, but only after nearest neighbourhood matching adjustment. No result reported in the first intervention or in the unmatched sample. Importantly, both interventions have no significant impact on any of the other measures of school investment and completion rates we employed in this study.

The results raise questions about the ability of citizens to hold representatives accountable using bottom-up strategies. Although studies like @bjorkman2009power, @fiala2017social and @reinikka2005fighting report better policy outcomes after providing information to local communities, our findings are closer to @banerjee2010pitfalls, @keefer2014mass, @lieberman2014does, @bjorkman2017experimental, @olken2007monitoring and @raffler2018weakness and offer mixed evidence that information interventions lead to greater government responsiveness. Our analysis suggests that grassroots monitoring does not affect government behaviour in Brazil, and other models of political accountability may be more effective for public service provision.

@@

Transparência Brasil partnered with the Brazilian branch of _Engineers without Borders_ (EWB), an independent non-governmental organisation[^ewb], to provide technical assessment of school completion rates. The EWB bases their evaluations on photos and GPS coordinates sent through the app. The engineers' reports are later uploaded to the TDP database and stored on the users' mobile devices so citizens can follow the progress of the constructions.

@@

Between August 2017 and August 2018, we implemented two interventions to measure the effect of the TDP app on six school construction outcomes. The outcomes are: 1) the percentage of public investment allocated to school constructions; 2) the difference between the percentage invested before and after the interventions; 3) the number of finished constructions; 4) the number of cancelled constructions; 5) the number of updated conclusion dates; 6) a placebo outcome indicating the percentage of the investment executed before the impact evaluation started. We expect the first five factors to increase and the last one to remain unchanged.

@@

Moreover, it is possible that a mobile app like TDP is simply not sufficient to promote collective action. @raffler2018weakness [29] point out several reasons why information interventions might lead to higher pressure on public agents, such as citizens' apathy towards their ability to make public agents accountable, lack of official channels to support popular demands, or that constituents have more pressing needs than monitoring representatives. Furthermore, individuals may also believe they are better-off by free-riding on other users of the TDP app, thus reaping the benefits of having faster school constructions without incurring personal costs [@olson1965logic].    
In sum, our experiments suggest that popular participation and bottom-up monitoring may not be effective to improve public service delivery. Nevertheless, the null findings are informative to researchers and policy-makers. The most important recommendation derived from this study is that interventions targeting elite groups, such as lobbyists or civil servants, may render better policy outcomes than those focused at the grassroots levels. Another core lesson is that although digital interventions are promising means to deliver information, perhaps they do not have the same impact as personal, face-to-face communication. Since many developing countries share Brazil's issues with education provision, the shortcomings we describe here serve as warnings for future interventions.

>>>>

One of the key determinants of efficient public services provision is a robust accountability system [@besley2003incentives; @cameron2004public; @ferejohn1986incumbent; @o1998horizontal]. In its standard definition, accountability is understood as the process of holding authorities responsible for their actions [@finer1941administrative; @mulgan2000accountability; @o1990bureaucratic]. Past studies show that accountability has a positive impact on governance, as it ensures that politicians act on behalf of voters [@freire2010ngp; @moncrieffe1998reconceptualizing], reduces the opportunities for rent-seeking and corruption [@deininger2005does; @wenar2006accountability] and improves the quality of public services [@adsera2003you; @bjorkman2009power]. Recent research also suggests that accountability leads to higher economic growth because it limits state discretion in the economy and increases long-term investments in human capital [@benhabib2010economic; @suebvises2018social; @ponzetto2018social].

@@

We use the TDP app to conduct two experimental interventions and test its impact on six outcomes related to school completion rates and complaints to public authorities. Overall, our results show that providing information to citizens has no consistent impact on policy outcomes. We find that the TDP app increased the likelihood of school completion by 7 percent in the first experiment, but the effect is null in the second intervention. We also find that the app increased the percentage of the executed budget by 157.5 percent in our second experiment, but only after nearest neighbourhood matching adjustment. Importantly, both interventions have no significant impact on any of the other measures of school investment and completion rates we employed in this study.

The results raise questions about the ability of citizens to hold representatives accountable using bottom-up strategies. Although studies like @bjorkman2009power, @fiala2017social and @reinikka2005fighting report better policy outcomes after providing information to local communities, our findings are closer to @banerjee2010pitfalls, @keefer2014mass, @lieberman2014does, @bjorkman2017experimental, @olken2007monitoring and @raffler2018weakness and offer little evidence that information interventions lead to greater government responsiveness. Our analysis suggests that grassroots monitoring does not affect government behaviour in Brazil, and other models of political accountability may be more effective for public service provision.

@@

Transparência Brasil partnered with the Brazilian branch of _Engineers without Borders_ (EWB), an independent non-governmental organisation[^ewb], to provide technical assessment of school completion rates based on user-submitted photos and GPS coordinates. The engineers' reports are later uploaded to the TDP database and stored on the users' mobile devices so citizens can follow the progress of the constructions.

@@

Between August 2017 and August 2018, we implemented two interventions to measure the effect of the TDP app on six school construction outcomes. The outcomes are: 1) a placebo outcome indicating the percentage of the investment executed before the impact evaluation started; 2) the percentage of public investment allocated to school constructions; 3) the difference between the percentage invested before and after the interventions; 4) the number of finished constructions; 5) the number of cancelled constructions; 6) the number of updated conclusion dates. We expect the first outcome to remain unchanged and the five remaining ones to increase.

@@

Moreover, it is possible that a mobile app like TDP is simply not sufficient to promote collective action. @raffler2018weakness [29] point out several reasons why information interventions does not lead to higher pressure on public agents, such as citizens' apathy towards their ability to make public agents accountable, lack of official channels to support popular demands, or that constituents have more pressing needs than monitoring representatives. Furthermore, individuals may also believe they are better-off by free-riding on other users of the TDP app, thus reaping the benefits of having faster school constructions without incurring personal costs [@olson1965logic].

In sum, our experiments suggest that popular participation and bottom-up monitoring may not be effective to improve public service delivery. Nevertheless, the null findings are informative to researchers and policy-makers. The most important recommendation derived from this study is that interventions targeting elite groups, such as lobbyists or civil servants, may render better policy outcomes than those focused at the grassroots levels. Another core lesson is that although digital interventions are promising means to deliver information, perhaps they do not have the same impact as personal, face-to-face communication. Since many developing countries share Brazil's issues with education provision, the shortcomings we describe here serve as warnings for future interventions.
